---
title: "Reproducible Research: Course Project 1"
output: html_document
---


Reading data, removing missing values and importing libraries
```{r, message=FALSE}
library(ggplot2)
require(gridExtra)

data <- read.csv('activity.csv')
```

## Plotting histogram of steps per day and calculating mean and median steps per day
```{r}
steps_per_day <-aggregate(data$steps, by=list(data$date), FUN=sum)
mean_steps_per_day <- mean(steps_per_day$x, na.rm = TRUE)
median_steps_per_day <- median(steps_per_day$x, na.rm = TRUE)

q <- qplot(steps_per_day$x, 
           main = 'Steps per day', 
           xlab = 'Number of steps per day', 
           geom = 'histogram', 
           binwidth = 2500,
           col = I("blue"), 
           fill = I("light blue"))
print(q)

```

Average number of steps per day: `r format(mean_steps_per_day, digits = 7)`

Median number of steps per day: `r format(median_steps_per_day, digits = 7)`

## Average number of steps per 5 min interval and handling missing values: 

Averaging number of steps every 5 min interval over all dates in the data set

```{r}


five_min_intervals <- aggregate(data$steps, by=list(data$interval), FUN=mean, na.rm=TRUE)
names(five_min_intervals) <- c("Interval", "Average number of steps")

```

Handling missing values by filling all occurances of NA in steps column with the average number of steps on the given 5 min interval, averaged over all dates in the data set
```{r}
missing_values <- sum(is.na(data))

for(i in unique(data$interval)){
  data_subsection <- data[data$interval == i, ]
  data_subsection$steps[is.na(data_subsection$steps)] <- floor(five_min_intervals$`Average number of steps`[five_min_intervals$Interval == i])
  data[data$interval == i, ] <- data_subsection
}


steps_per_day <-aggregate(data$steps, by=list(data$date), FUN=sum)
mean_steps_per_day <- mean(steps_per_day$x, na.rm = TRUE)
median_steps_per_day <- median(steps_per_day$x, na.rm = TRUE)

q <- qplot(steps_per_day$x, 
           main = 'Steps per day', 
           xlab = 'Number of steps per day', 
           geom = 'histogram', 
           binwidth = 2500,
           col = I("blue"), 
           fill = I("light blue"))
print(q)
```

There are `r missing_values` rows with missing data in the dataset. 
The new mean and median of the data set with the missing values replaced by averages are `r  format(mean_steps_per_day, digits = 7)` and `r format(median_steps_per_day, digits = 7)` respectively. The mean value is more or less the same as before, slightly lower, but the median has sunk by about a 100. It is clear from the two histogram that the one produced from the data set with the replaced missing values are sharper, it peaks higher.


Plotting a line plot over of the average number of steps in every 5 min interval (y-axis) agains the interval number (x-axis)
```{r}
q1 <- qplot(five_min_intervals$Interval, five_min_intervals$`Average number of steps`, 
            xlab = 'Interval number', 
            ylab = 'Number of steps', 
            main = 'Averaged number of steps in each 5 min interval', 
            geom  = 'line')
print(q1)

interval_with_max_numer_of_steps <- max(five_min_intervals)
```

Interval with the on average highest number of steps: `r format(interval_with_max_numer_of_steps, digits = 7)`


## Average number of steps per 5 minute interval during weekdays
```{r}
data$weekend = chron::is.weekend(data$date)


weekday <- data[!data$weekend, ]
weekday <- aggregate(weekday$steps, by=list(weekday$interval), FUN=mean)
names(weekday) <- c("Interval", "Average number of steps")

q2 <- qplot(weekday$Interval, weekday$`Average number of steps`, 
            xlab = 'Interval number', 
            ylab = 'Number of steps', 
            main = 'Averaged number of steps in each 5 min interval - Weekdays', 
            geom = 'line')

weekend <- data[data$weekend, ]
weekend <- aggregate(weekend$steps, by=list(weekend$interval), FUN=mean)
names(weekend) <- c("Interval", "Average number of steps")

q3 <- qplot(weekend$Interval, weekend$`Average number of steps`, 
            xlab='Interval number', 
            ylab = 'Number of steps', 
            main = 'Averaged number of steps in each 5 min interval - Weekends', 
            geom = 'line')

q4 <- grid.arrange(q2, q3)
```

The activity pattern during weekends appear to be much more spread out across the day than during the weekdays. Weekdays have much more of a clear peak around the 800th 5-min interval
